---
title: 'HW #6 Untidy Data'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r load-pkgs, eval=TRUE}
library('dplyr')
library('tidyr')
library("stringr")
library("knitr")
library("psych")
```
1. Load Data Frame from website
```{r load-text-by-line, eval=TRUE}
l <- read.csv(
              "/Users/scottkarr/IS607Spring2016/hw6/more/untidy-data.csv",
              sep=",",
              na.strings = "",
              blank.lines.skip = TRUE,
              col.names = c("Airline", "Time",  "City1", "City2", "City3", "City4","City5"),
              stringsAsFactors=FALSE
    )
df = data.frame(l)
kable(head(df), align = 'l')
```


2.  Remove Null Row
```{r remove-null-row,eval=TRUE}
df_staging <- df[is.na(df$Airline) %in% FALSE | is.na(df$Time) %in% FALSE, ]
kable(head(df_staging))
```

3.  Fill Missing Airline data
```{r fill-missing-col,eval=TRUE}
# adds an Airline lag column
df_staging <- mutate(df_staging, lag = lag(Airline))
# indexes Airline missing data and assigns from lag column
df_staging$Airline[is.na(df_staging$Airline) == TRUE] <- df_staging$lag[is.na(df_staging$Airline) == TRUE]
kable(head(df_staging), align = 'l')
```

4.  Spread and combine city data
    *Note, gather may be more effective here but not on the first go-round
```{r tidy,eval=TRUE}
# Keep it simple, process 1 city at a time
df_city1 <- df_staging[,1:3]
df_city2 <- df_staging[,c(1:2,4)]
df_city3 <- df_staging[,c(1:2,5)]
df_city4 <- df_staging[,c(1:2,6)]
df_city5 <- df_staging[,c(1:2,7)]

# De-normalize the repeating groups
df_city1 <- spread(df_city1, Time, City1)
df_city1["City"] <- "City1"
df_city2 <- spread(df_city2, Time, City2)
df_city2["City"] <- "City2"
df_city3 <- spread(df_city3, Time, City3)
df_city3["City"] <- "City3"
df_city4 <- spread(df_city4, Time, City4)
df_city4["City"] <- "City4"
df_city5 <- spread(df_city5, Time, City5)
df_city5["City"] <- "City5"

# Combine longform for output
df_Output <- rbind(df_city1,df_city2,df_city3,df_city4,df_city5)
colnames(df_Output)[3] <- "OnTime"
df_Output <- df_Output[c(1,4,2,3)]

# Subset numeric columns to compute statistics
kable(df_Output, align = 'l')
kable(df_staging, align = 'l')
```

5. OK, so now that we have our arrival/departure data in long form, let's compute summary
statistics for on-time arrivals to see how the cities compare . . .

 * add on time percentage column so we're comparing apples with apples
 * show summary statistics for on time percentage as well as nominal delay and on-time counts
 * use boxplot to show on time percentages by city comparitively

```{r compute-stats,eval=TRUE}
df_Output <- mutate(df_Output, OnTimePct = (delayed/(delayed + OnTime)) * 100 )
nums <- sapply(df_Output, is.numeric)
df_stats <- df_Output[ , nums]
df_summary <- describe(df_stats)
kable(df_summary,align='l')
boxplot(split(df_Output$OnTimePct,df_Output$City),main='On Time Arrivals By City')
```
